Low Dimensional Embedding of Climate Data for Radio Astronomical Site Testing in the Colombian Andes

### 2017-06-02 pasp/paperclimate.tex
ultimas correciones oleo
matpart

\author{Germ¬\'an Chaparro Molano$^{1}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\ead{gchaparrom@ecci.edu.co}
\author{Oscar Leonardo Ram\'irez Su\'arez$^{1}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\ead{oramirezs@ecci.edu.co}
\author{Oscar Restrepo$^{1,2}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\address{$^2$Radio Astronomy Instrumentation Group, Universidad de Chile, Santiago de Chile, Chile}
\ead{orestrepog@ecci.edu.co}
\author{Alexander Mart\'inez$^{1,3}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\address{$^3$Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales, Bogot\'a, Colombia}
\vspace{10pt}
\begin{indented}
\item[]December 2016
\end{indented}

\begin{abstract}
This document describes the  preparation of an article using \LaTeXe\ and 
\verb"iopart.cls" (the IOP Publishing \LaTeXe\ preprint class file).
This class file is designed to help 
authors produce preprints in a form suitable for submission to any of the
journals listed in table~\ref{jlab1} on the next page.  You are not obliged to use this class file---we accept
submissions using all common \LaTeX\ class and style files.  The \verb"iopart.cls"
class file is supplied merely as a convenience for those authors who find it useful.
This document gives both general advice that applies whatever class file you use, and specific advice
that applies if you choose to use \verb"iopart.cls".

We also accept submissions in Word format.  Instructions for Word submissions are available via the `Author Guidelines' link at \verb"http://authors.iop.org".

If you have any queries about this document or any aspect of preparing your article for submission please contact us at the e-mail address given above.
\end{abstract}

% Uncomment for PACS numbers
%\pacs{00.00, 20.00, 42.10}
%
% Uncomment for keywords
%\vspace{2pc}
\noindent{\it Keywords}: Water Vapor, Atmospheric Effects, Site Testing, Radio Astronomy, Satellite Communication, \\
%
% Uncomment for Submitted to journal title message
\submitto{PASP}
%
% Uncomment if a separate title page is required
%\maketitle
% 
% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
%\ioptwocol
%

@@

The development of astronomical instrumentation technology in the 0.2-2 THz range has been rapidly growing in recent years. Another recent surge in interest in telecommunications THz technology has appeared recently due to the saturation of the ``classical radio window'' in telecommunications  \cite{newera}. For this reason, there has been development of atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal}. Near-equatorial countries face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}. In northern South America, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}. However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the northern Andes could be suitable candidates for a world-class millimeter-wave observatory. \\

@@

The meteorological data used in this study were gathered over 30 years (1981-2010) in 2046 weather stations monitored by the Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales (IDEAM) in Colombia (Figure \ref{totmap}). The  variables that we considered relevant to our work were: Elevation (in meters above sea level, or masl), Precipitation (mm/mo), Rain Days (d/mo),  Relative Humidity (\%), and Sunshine Duration (h/d). The climatological variables are reported as multi-annual monthly data, i.e. each variable is reported monthly from January to December averaged over the 30 year range. For precipitation, each monthly datum corresponds to the cumulative monthly value.  For rain days, each monthly datum corresponds to the number of rain days for that month.  For relative humidity, each monthly datum corresponds to the  average daily value averaged over each month. For sunshine duration, each monthly datum corresponds to the daily value averaged over each month. All stations reported precipitation values, although only 2002 reported rain days, 445 reported relative humidity, and 336 reported sunshine duration. This means that not all stations registered all the meteorological variables relevant to this work. Thus, we classify stations according to which variables they measured (Table \ref{tabstype}).\\

@@

\begin{table}
\caption{\label{tabclu}Climate clusters according to a lowest BIC-based selection of Gaussian Mixture Models following Figure \ref{bic}. The selected clusters column refers to the climate clusters shown in Figures \ref{clusth} to \ref{clustd}. }
\begin{indented}
\item[]\begin{tabular}{@{}lclll}
\br
Variable&Symbol&Lowest BIC&Lowest BIC&Selected Clusters\\
&&No. of Clusters&Cov. Method &\\
\mr
Relative Humidity&$H$&2&Full&2\\
Precipitation &$R$&11&Full&2,3,7,11\\
Rain Days&$D$&6&Full&1,4,6\\
Sunshine Duration&$S$&5&Tied&1,4,5\\
\br
\end{tabular}
\end{indented}
\end{table}

@@

Thus, the probability that a station $j=(0,1,...,N_\mathrm{stations})$ of type $T_i$, meets all four criteria ($C$) and appears in our shortlist ($A$) given that it is located at an elevation $h$, can be written as $P_j=P(C\cap A\cap T_i\ |\ h)$. We cannot  calculate exactly this probability, as it depends on unknown unknowns, i.e. whether a station that did not measure a given variable in reality belongs to one of our preferred climatological clusters for that variable (Table \ref{tabclu}). Thus we will use Bayes' theorem to help estimate this probability using simpler conditional probabilities. First we control for the inhomogeneous distribution of elevations $P(h)$,
\begin{equation}
P_j=P(C\cap A\cap T_i\ |\ h)=\frac{P(C\cap A\cap T_i\cap h)}{P(h)}\ .
\end{equation}
The joint probability of all events $C\cap A\cap T_i\cap h$ can be rewritten as,
\begin{equation}
P_j=\frac{P(h\ |\ C\cap A\cap T_i)}{P(h)}P(C\cap A\cap T_i)\ .
\end{equation}
If a station satisfies all four criteria, $P(C\cap A\cap T_i)=1$. This probability can be expressed in terms of simpler conditional probabilities,
\begin{equation}\label{peejay}
P_j=\frac{P(h\ |\ C\cap A\cap T_i)}{P(h)}P(C\ |\ A\cap T_i)P(A\ |\ T_i)P(T_i)\ ,
\end{equation}
where 
\begin{itemize}
\item $P(h\ |\ C\cap A\cap T_i)$ is the distribution of elevations for a station type ($T_i$) on our shortlist ($C\cap A$), 
\item $P(A\ |\ T_i)$ is the probability that a station of type $T_i$ is in our shortlist, 
\item $P(T_i)$ is the probability that a station type is $T_i$, and 
\item $P(C\ |\ A\cap T_i)$ is the probability that a station meets all four criteria ($C$) given that it is in our shortlist ($A$) and is of type $T_i$. 
\end{itemize}
All of these probabilities except for the last one can be computed directly from the data\footnote{Distributions of elevation-related probabilities were approximated discretely using a bin size estimated from the mean elevation difference between stations when sorted by elevation. }.\\

 The probability $P(C\ |\ A\cap T_i)$ has to be estimated indirectly, as we cannot presume to know under which conditions a station can satisfy all our criteria. However, we can rewrite this probability as,
\begin{eqnarray}
P(C\ |\ A\cap T_i)&=P(R\cap D\cap H\cap S\ |\ A\cap T_i)\ , \\
\label{pclanti}
&=\frac{P(R\cap D\cap H\cap S\cap A\cap T_i)}{P(A\cap T_i)}\ . 
\end{eqnarray}
In order to estimate $P(C\ |\ A\cap T_i)$ we assume that the probability of a station meeting our criteria for a given number of variables is independent of the station type. To illustrate this, let us assume that a given station is of type 1, so it only measured precipitation and rain days, i.e.,
\begin{equation}
T_1=M_R\cap M_D\cap\neg M_H\cap\neg M_S\ .
\end{equation}
If this station of type $T_1$ is in our shortlist ($A$), the $R$ and $D$ criteria are met. Thus,
\begin{equation}
P(R\cap D\cap H\cap S\cap A\cap T_1)=P( H\cap S\ |\ A\cap T_1)\ .
\end{equation}
Substituting this into Eq. (\ref{pclanti}),
\begin{equation}
P(C\ |\ A\cap T_1)=\frac{P( H\cap S\ |\ A\cap T_1)}{P(A\cap T_1)}=P(H\cap S\ |\ A\cap T_1) \ .
\end{equation}
The last term in the previous expression is unknown, but we can approximate it assuming, as mentioned above, that it is independent of the station type. Therefore, 
\begin{equation}
P(C\ |\ A\cap T_)\simeq P(H\cap S)\ .
\end{equation}
The probability on the right hand side of the previous equation can be directly computed from the data, as it only requires counting how many stations meet both $H$ and $S$ criteria. This procedure can be extended to all other station types.\\

@@

\begin{table}
\caption{\label{tabfin}Stations in two regions of interest (RI) showing a high-altitude daytime dry, clear-sky climate, organized by location, Bayesian probability quality measure ($Q_j$), distance to cluster center ($r_c$), mean precipitation  ($\bar{R}$), mean rain days ($\bar{D}$), mean relative humidity  ($\bar{H}$), mean sunshine duration  ($\bar{S}$), and altitude ($h$). Detailed table (including all clusters) available at \texttt{https://github.com/saint-germain/ideam} \\}
%https://tex.stackexchange.com/questions/98388/how-to-make-table-with-rotated-table-headers-in-latex

\begin{tabular}{clrrrrrrr}
\toprule
RI&                   Municipality  &  $Q_j$ &  $r_c$  &   $\bar{R}$ &   $\bar{D}$  &   $\bar{H}$  &  $\bar{S}$ &  $h$ \\
 &                   &    &    (km) &   (mm/mo) &  (d/mo) &   (\%) &   (h/d) &  (masl) \\
\midrule
            &                       Nemocon &            2.2 &                        28 &                        66 &                   ND &                        ND &                          ND &    2600 \\
            &                          Susa &            4.6 &                        18 &                        83 &                   12 &                        ND &                          ND &    2600 \\
            &                      Cucunuba &            4.7 &                         6 &                        57 &                    9 &                        ND &                          ND &    2620 \\
            &                     Sutatausa &            4.8 &                         6 &                        57 &                   10 &                        ND &                          ND &    2700 \\
            &                      Guacheta &            4.9 &                        16 &                        73 &                   10 &                        ND &                          ND &    2690 \\
            &                   Lenguazaque &            5.2 &                        10 &                        63 &                   12 &                        ND &                          ND &    2650 \\
            &                      Simijaca &            5.2 &                        23 &                        67 &                   11 &                        ND &                          ND &    2572 \\
            &                        Suesca &            5.6 &                        20 &                        59 &                   10 &                        ND &                          ND &    2575 \\
                      \rotatebox{90}{\rlap{Valle de Ubat\'e}}
            &              Carmen de Carupa &            6.1 &                        12 &                        54 &                   12 &                        ND &                          ND &    2970 \\
            &                      Cucunuba &            6.1 &                         7 &                        51 &                    5 &                        ND &                          ND &    2562 \\
            &                   Lenguazaque &            6.1 &                         9 &                        57 &                   10 &                        ND &                          ND &    2585 \\
            &                          Susa &            6.1 &                        11 &                        60 &                    8 &                        ND &                          ND &    3130 \\
            &   Ubate &            6.1 &                         4 &                        62 &                   13 &                        ND &                          ND &    2555 \\
            &   Ubate &            6.1 &                         5 &                        62 &                   12 &                        ND &                          ND &    2610 \\
            &                       Nemocon &            7.6 &                        20 &                        52 &                   10 &                        77 &                           4 &    2580 \\
                \midrule
            &                       Duitama &            4.0 &                        17 &                        69 &                   10 &                        ND &                          ND &    2590 \\
            &                         Paipa &            4.0 &                        12 &                        74 &                    9 &                        ND &                          ND &    2900 \\
            &                         Nobsa &            5.0 &                         6 &                        66 &                   10 &                        ND &                          ND &    2500 \\
            &         Santa Rosa &            5.0 &                        10 &                        70 &                    9 &                        ND &                          ND &    2500 \\
            &                      Sogamoso &            5.0 &                         4 &                        59 &                   10 &                        ND &                          ND &    2500 \\
            &                      Sogamoso &            5.0 &                         7 &                        60 &                   12 &                        ND &                          ND &    2500 \\
            &                      Tibasosa &            5.0 &                         5 &                        62 &                    7 &                        ND &                          ND &    2500 \\
                                              \rotatebox{90}{\rlap{~Valle del Sol}}
            &                           Iza &            5.6 &                        12 &                        52 &                    9 &                        ND &                          ND &    2470 \\
            &                    Firavitoba &            6.1 &                         6 &                        57 &                    9 &                        ND &                          ND &    2486 \\
            &                        Mongui &            6.1 &                        12 &                        65 &                    9 &                        ND &                          ND &    2970 \\
            &                      Sogamoso &            6.1 &                        11 &                        65 &                   10 &                        ND &                          ND &    3225 \\
            &                      Sogamoso &            6.7 &                         5 &                        61 &                   12 &                        75 &                           5 &    2500 \\
            &                         Nobsa &           10.0 &                        10 &                        68 &                   13 &                        74 &                           4 &    2530 \\

@@

In this paper we seek to find plausible locations for a high-mountain mm-wave astronomical observatory in the northern Andes of Colombia, by analizing 30 years of climate data from 2046 weather stations. By low-dimensionally embedding the data, we are able to group together and correlate climate behavior indicative of daytime dry, clear-sky weather conditions at elevations higher than 2000 masl.  A repository with all our code and the full dataset is available at \texttt{https://github.com/saint-germain/ideam}.\\

From our shortlist of 79 stations that met our criteria, we selected 70 stations located  in the Cundinamarca-Boyac\'a \emph{altiplano} region of Colombia, which is a high-mountain basin with an average altitude of 2600 masl. Weather stations in our shortlist but outside this region are too sparsely located to suggest a candidate region of interest. \\

@@

We summarize the information about two regions of interest (Valle de Ubat\'e and Valle del Sol) in Table \ref{tabfin}, selected based on a Bayesian probabilistic quality index. Most of the stations in this table are located at elevations above 2600 masl and below 3000 masl. We validated our results using satellite measurements of the upper bound for precipitable water vapor \cite{suen2014,suen2016} for those two regions, whis is reported to be $\sim15$ mm. This means that it is not implausible to find a site with an even lower water vapor column in one of the regions of interest identified here. Some mountains nearby Valle de Ubat\'e and Valle del Sol  have altitudes of up to 3400 masl, and therefore warrant further measurements using radio sondes and radiometers. \\

The Villa de Leyva region, despite having a dry, sunny climate, is located at a comparatively low altitude (2200 masl), which means that the precipitable water vapor column above this region is too high ($>20$ mm according to the \citeasnoun{suen2016} maps).\\

>>>>

\author{Germ¬\'an Chaparro Molano$^{1}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\ead{gchaparrom@ecci.edu.co}
\author{Oscar Leonardo Ram\'irez Su\'arez$^{1}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\ead{oramirezs@ecci.edu.co}
\author{Oscar Restrepo$^{1,2}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\address{$^2$Radio Astronomy Instrumentation Group, Universidad de Chile, Santiago de Chile, Chile}
\ead{orestrepog@ecci.edu.co}
\author{Alexander Mart\'inez$^{1,3}$}
\address{$^1$Grupo de Simulaci\'on, An\'alisis y Modelado, Vicerrector\'ia de Investigaci\'on, Universidad ECCI, Bogot\'a, Colombia}
\address{$^3$Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales, Bogot\'a, Colombia}
\vspace{10pt}
\begin{indented}
\item[]
\end{indented}

\begin{abstract}
We set out to evaluate the potential of the Colombian Andes for millimeter-wave astronomical observations. Previous studies for astronomical site testing in this region have suggested that nighttime humidity and cloud cover conditions make most sites unsuitable for professional visible-light observations. Millimeter observations can be done during the day, but require that the precipitable water vapor column above a site stays below $\sim$10 mm. Due to a lack of direct radiometric or radiosonde measurements, we present a method for correlating climate data from weather stations to sites with a low precipitable water vapor column. We use unsupervised learning techniques to low-dimensionally embed climate data (precipitation, rain days, relative humidity, and sunshine duration) in order to group together stations with similar long-term climate behavior. The data were taken over a period of 30 years by 2046 weather stations across the Colombian territory. We find 6 regions with unusually dry, clear-sky conditions, ranging in elevations from 2200 to 3800 masl. We evaluate the suitability of each region using a quality index derived from a Bayesian probabilistic analysis of the station type and elevation distributions. Two of these regions show a high probability of having an exceptionally low precipitable water vapor column. We compared our results with global precipitable water vapor maps and find a plausible geographical correlation with regions with low water vapor columns ($\sim15$ mm) at an accuracy of $\sim20$ km. Our methods can be applied to similar datasets taken in other countries as a first step toward astronomical site evaluation.
\end{abstract}

% Uncomment for PACS numbers
%\pacs{00.00, 20.00, 42.10}
%
% Uncomment for keywords
%\vspace{2pc}
\noindent{\it Keywords}: atmospheric effects -- methods: data analysis -- methods: statistical -- site testing \\
%
% Uncomment for Submitted to journal title message
\submitto{PASP}
%
% Uncomment if a separate title page is required
%\maketitle
% 
% For two-column output uncomment the next line and choose [10pt] rather than [12pt] in the \documentclass declaration
%\ioptwocol
%

@@

The development of astronomical instrumentation technology in the 0.1-2 THz range has been rapidly growing in recent years. Another recent surge in interest in telecommunications THz technology has appeared recently due to the saturation of the ``classical radio window'' in telecommunications  \cite{newera}. For this reason, atmospheric models for absorption of THz photons \cite{rosenkranz,lababs} and artifact removal models \cite{removal} have been developed. Countries near the Equator face a challenge in using this frequency band for astronomical observations and telecommunications using terabit satellite links  \cite{suen2016} due to the presence of a tropical belt of dense water vapor which efficiently absorbs THz radiation \cite{tropicalbelt}. In northern South America, previous studies of astronomical site testing in the visible range have shown that high nighttime humidity conditions make this region suitable only for educational observatories \cite{pinzon}. However, considering that millimeter/sub-millimeter observations need not be done during the night, unusually dry, clear-daytime-sky, high altitude regions in the northern Andes could be suitable candidates for a world-class millimeter-wave observatory. \\

@@

The meteorological data used in this study were gathered over 30 years (1981-2010) in 2046 weather stations monitored by the Instituto de Hidrolog\'ia, Meteorolog\'ia y Estudios Ambientales (IDEAM) in Colombia (Figure \ref{totmap}). The  variables that we considered relevant to our work were: Elevation (in meters above sea level, or masl), Precipitation (mm/mo), Rain Days (d/mo),  Relative Humidity (\%), and Sunshine Duration (h/d). The climatological variables are reported as multi-annual monthly data, i.e. each variable is reported monthly from January to December averaged over the 30 year range. For precipitation, each monthly datum corresponds to the cumulative monthly value.  For rain days, each monthly datum corresponds to the number of rain days for that month.  For relative humidity, each monthly datum corresponds to the  average daily value averaged over each month. For sunshine duration, each monthly datum corresponds to the daily value averaged over each month. All stations reported precipitation values, although only 2002 reported rain days, 445 reported relative humidity, and 336 reported sunshine duration. This means that not all stations registered all the meteorological variables relevant to this work. Thus, we classify stations according to which variables they measured (Table \ref{tabstype}). The data is public and can be found in the repository for this paper at \texttt{https://github.com/saint-germain/ideam} .\\

@@

\begin{table}
\caption{\label{tabclu}Climate clusters according to a lowest BIC-based selection of Gaussian Mixture Models following Figure \ref{bic}. The selected clusters column refers to the climate clusters shown in Figures \ref{clusth} to \ref{clustd}. }
\begin{indented}
\item[]\begin{tabular}{@{}lclll}
\br
Variable&Symbol& No. of&Covariance&Selected \\
&&clusters&Method &Clusters\\
\mr
Relative Humidity&$H$&2&Full&2\\
Precipitation &$R$&11&Full&2,3,7,11\\
Rain Days&$D$&6&Full&1,4,6\\
Sunshine Duration&$S$&5&Tied&1,4,5\\
\br
\end{tabular}
\end{indented}
\end{table}

@@

Thus, the probability that a station $j=\{0,1,...,N_\mathrm{shortlist}\}$ of type $T_i$, meets all four criteria ($C$) and appears in our shortlist ($A$) given that it is located at an elevation $h$, can be written as $P_j=P(C\cap A\cap T_i\ |\ h)$. We cannot  calculate exactly this probability, as it depends on unknown unknowns, i.e. on whether a station that did not measure a given variable in reality belongs to one of our preferred climatological clusters for that variable (Table \ref{tabclu}). Thus we will use Bayes' theorem to help estimate this probability using simpler conditional probabilities. First we control for the inhomogeneous distribution of elevations $P(h)$,
\begin{equation}
P_j=P(C\cap A\cap T_i\ |\ h)=\frac{P(C\cap A\cap T_i\cap h)}{P(h)}\ .
\end{equation}
The joint probability of all events $C\cap A\cap T_i\cap h$ can be rewritten as,
\begin{equation}
P_j=\frac{P(h\ |\ C\cap A\cap T_i)}{P(h)}P(C\cap A\cap T_i)\ .
\end{equation}
If a station satisfies all four criteria, $P(C\cap A\cap T_i)=1$. This probability can be expressed in terms of simpler conditional probabilities,
\begin{equation}\label{peejay}
P_j=\frac{P(h\ |\ C\cap A\cap T_i)}{P(h)}P(C\ |\ A\cap T_i)P(A\ |\ T_i)P(T_i)\ ,
\end{equation}
where 
\begin{itemize}
\item $P(h\ |\ C\cap A\cap T_i)$ is the distribution of elevations for a station type ($T_i$) on our shortlist ($C\cap A$), 
\item $P(A\ |\ T_i)$ is the probability that a station of type $T_i$ is in our shortlist, 
\item $P(T_i)$ is the probability that a station type is $T_i$, and 
\item $P(C\ |\ A\cap T_i)$ is the probability that a station meets all four criteria ($C$) given that it is in our shortlist ($A$) and is of type $T_i$. 
\end{itemize}
All of these probabilities except for the last one can be computed directly from the data\footnote{Distributions of elevation-related probabilities were approximated discretely using a bin size estimated from the mean elevation difference between stations when sorted by elevation. }.\\

\begin{equation}\label{pclanti}
P(C\ |\ A\cap T_i)=\frac{P(C\cap A\cap T_i)}{P(A\cap T_i)}\  .
\end{equation}
In order to estimate $P(C\ |\ A\cap T_i)$ we assume that the probability of a station meeting our criteria for a given number of variables is independent of the station type. To illustrate this, let us assume that a given station is of type 1, so it only measured precipitation and rain days, i.e.,
\begin{equation}
T_1=M_R\cap M_D\cap\neg M_H\cap\neg M_S\ .
\end{equation}
If this station of type $T_1$ is in our shortlist ($A$), the $R$ and $D$ criteria are already met, so $A\cap R \cap D=A$. Thus,
\begin{equation}
P(C\cap A\cap T_1)=P( H\cap S\ |\ A\cap T_1)\ .
\end{equation}
Substituting this into Eq. (\ref{pclanti}),
\begin{equation}
P(C\ |\ A\cap T_1)=\frac{P( H\cap S\ |\ A\cap T_1)}{P(A\cap T_1)}=P(H\cap S\ |\ A\cap T_1) \ .
\end{equation}
The last term in the previous expression is unknown, but we can approximate it by assuming that it is independent of the station type. Therefore, 
\begin{equation}
P(C\ |\ A\cap T_1)\simeq P(H\cap S)\ .
\end{equation}
The probability on the right hand side of the previous equation can be directly computed from the data, as it only requires counting how many stations meet both $H$ and $S$ criteria. This procedure can be extended to all other station types.\\

@@

\begin{table}
\caption{\label{tabfin}Stations in two regions of interest showing a high-elevation daytime dry, clear-sky climate, organized by location, Bayesian probability quality measure ($Q_j$), distance to cluster center ($r_c$), mean precipitation  ($\bar{R}$), mean rain days ($\bar{D}$), mean relative humidity  ($\bar{H}$), mean sunshine duration  ($\bar{S}$), and elevation ($h$). Detailed table (including all clusters) available at \texttt{https://github.com/saint-germain/ideam} .}

\begin{tabular}{cllccccccc}
\toprule
Region of&   Station&                Municipality  &  $Q_j$ &  $r_c$  &   $\bar{R}$ &   $\bar{D}$  &   $\bar{H}$  &  $\bar{S}$ &  $h$ \\
interest &    code   &            &    &    (km) &   (mm/mo) &  (d/mo) &   (\%) &   (h/d) &  (masl) \\
\midrule
  &   21201410 &           Nemocon &  2.2 &            28 &       66 &       ND &    ND &  ND &    2600 \\
  &   24011060 &              Susa &  4.6 &            18 &       83 &       12 &    ND &  ND &    2600 \\
  &   24010140 &          Cucunuba &  4.7 &             6 &       57 &        9 &    ND &  ND &    2620 \\
  &   24015210 &         Sutatausa &  4.8 &             6 &       57 &       10 &    ND &  ND &    2700 \\
  &   24010170 &          Guacheta &  4.9 &            16 &       73 &       10 &    ND &  ND &    2690 \\
  &   24010070 &       Lenguazaque &  5.2 &            10 &       63 &       12 &    ND &  ND &    2650 \\
  &   24015130 &          Simijaca &  5.2 &            23 &       67 &       11 &    ND &  ND &    2572 \\
  &   21201620 &            Suesca &  5.6 &            20 &       59 &       10 &    ND &  ND &    2575 \\
                                       \rotatebox{90}{\rlap{Valle de Ubat\'e}}
  &   24010610 &  Carmen de Carupa &  6.1 &            12 &       54 &       12 &    ND &  ND &    2970 \\
  &   24011080 &          Cucunuba &  6.1 &             7 &       51 &        5 &    ND &  ND &    2562 \\
  &   24010280 &       Lenguazaque &  6.1 &             9 &       57 &       10 &    ND &  ND &    2585 \\
  &   24010440 &              Susa &  6.1 &            11 &       60 &        8 &    ND &  ND &    3130 \\
  &   24011090 &             Ubate &  6.1 &             4 &       62 &       13 &    ND &  ND &    2555 \\
  &   24015110 &             Ubate &  6.1 &             5 &       62 &       12 &    ND &  ND &    2610 \\
  &   21205400 &           Nemocon &  7.6 &            20 &       52 &       10 &    77 &   4 &    2580 \\
      \midrule
  &   24030760 &           Duitama &  4.0 &            17 &       69 &       10 &    ND &  ND &    2590 \\
  &   24030510 &             Paipa &  4.0 &            12 &       74 &        9 &    ND &  ND &    2900 \\
  &   24030790 &             Nobsa &  5.0 &             6 &       66 &       10 &    ND &  ND &    2500 \\
  &   24031040 &        Santa Rosa &  5.0 &            10 &       70 &        9 &    ND &  ND &    2500 \\
  &   24035140 &          Sogamoso &  5.0 &             4 &       59 &       10 &    ND &  ND &    2500 \\
  &   24030940 &          Sogamoso &  5.0 &             7 &       60 &       12 &    ND &  ND &    2500 \\
  &   24030410 &          Tibasosa &  5.0 &             5 &       62 &        7 &    ND &  ND &    2500 \\
                                      \rotatebox{90}{\rlap{~Valle del Sol}}
  &   24030410 &               Iza &  5.6 &            12 &       52 &        9 &    ND &  ND &    2470 \\
  &   24030540 &        Firavitoba &  6.1 &             6 &       57 &        9 &    ND &  ND &    2486 \\
  &   24030190 &            Mongui &  6.1 &            12 &       65 &        9 &    ND &  ND &    2970 \\
  &   24030760 &          Sogamoso &  6.1 &            11 &       65 &       10 &    ND &  ND &    3225 \\
  &   24035340 &          Sogamoso &  6.7 &             5 &       61 &       12 &    75 &   5 &    2500 \\
  &   24035150 &             Nobsa & 10.0 &            10 &       68 &       13 &    74 &   4 &    2530 \\

@@

In this paper we find plausible locations for a high-mountain mm-wave astronomical observatory in the northern Andes of Colombia, by analizing 30 years of climate data from 2046 weather stations. By low-dimensionally embedding the data, we are able to group together and correlate climate behavior indicative of daytime dry, clear-sky weather conditions.  A repository with all our code and the full dataset is available at \texttt{https://github.com/saint-germain/ideam}.\\

From our shortlist of 79 stations at elevations higher than 2000 masl that met our criteria, we selected 70 stations located  in the Cundinamarca-Boyac\'a \emph{altiplano} region of Colombia, which is a high-mountain basin with an average elevation of 2600 masl. Weather stations in our shortlist but outside this region are too sparsely located to suggest a candidate region of interest. \\

@@

We summarize the information about two regions of interest (Valle de Ubat\'e and Valle del Sol) in Table \ref{tabfin}, selected based on a Bayesian probabilistic quality index. Most of the stations in this table are located at elevations above 2600 masl and below 3000 masl. We validated our results using satellite measurements of the upper bound for mean precipitable water vapor \cite{suen2014,suen2016} for those two regions, which is reported to be $\sim15$ mm in $\sim20$ km resolution maps. This means that it is not implausible to find a site with an even lower water vapor column in one of the regions of interest identified here. Some mountains nearby Valle de Ubat\'e and Valle del Sol  have elevations of up to 3400 masl, and therefore warrant further measurements using radio sondes and radiometers. \\

The Villa de Leyva region, despite having a dry, sunny climate, is located at a comparatively low elevation (2200 masl), which means that the precipitable water vapor column above this region is too high ($>20$ mm according to the \cite{suen2016} maps).\\
